On the Synchronization Power of the OpenFlow Data Plane

### 2015-05-22 inband-sync.tex
Sec 2-4

\newcommand{\hide}[1]{}
\newcommand{\concat}[0]{\oplus}
\newcommand{\append}[0]{\frown}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\cas}{CAS\xspace}
\newcommand{\claimcheck}{check\xspace}
\newcommand{\compare}{compare\xspace}
\newcommand{\memwrite}{write\xspace}

@@

%\title{Solving Consensus with OPenFlow}

@@

%[[PK sorry for stubborness :)
\title{On the Synchronization Power of the OPenFlow Data Plane}
%\\ {\Large Towards Atomic Transactions In-Band}}
%]]

@@

We show that powerful synchronization constructs can be implemented
directly \emph{in OpenFlow}, i.e., using the \emoh{de facto} standard SDN
protocol. 
In our implementation, a part of the data-plane configuration space is
used as a shared memory storing information about  contention and
conflicts between distributed controllers.
A transaction can then contain standard OpenFlow control operation
as well as  \emph{synchronization primitives} operating on this shared
memory. 
Our synchronization primitives allow the controllers to define complex notions
of conflicts that can be used for coordinating configuration updates that
operate over \emph{independent flow spaces}, but where dependencies are introduced
in the mapping of the flows to the physical infrastructure. This is
important, e.g., for load-balancing applications.

@@

More precisely, the match field of a rule is 
a ternary pattern over packet header fields
(a value and a mask).
\petr{example?}
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
\petr{unclear what ``sometimes'' means here}
%(For instance, the meta-data field is maskable.)

A set of rules, associated with \emph{priority} levels, form \emph{flow
  entries} in a \emph{flow table}, and the 
entries are ordered according to their priorities.
Furthermore, an OpenFlow switch may maintain multiple flow tables
forming a \emph{pipeline}.
Data-plane packets are checked against flow entries in the pipelined
tables to determine the set of actions the switch should perform on the packet.      
A data-plane packet arriving at a switch is first checked against
the rule with the highest priority and in table $0$. 
If the header of the packet fits the match fields of that rule, 
a default \emph{instruction} associates the packet the corresponding action.
Otherwise, the packet is checked against flow  entries with lower
priorities.  
Specific instructions can be used to define
additional tables to be visited (\textit{goto} instructions), to
modify the set of actions associated with the packet (by adding,
deleting or modifying actions), or immediately apply some actions in the action set. 
During the traversal of the flow tables, the packet can also be equipped with \emph{meta-tags} to
transfer temporary information between tables.
The meta-data field is part of the header that can be
inserted or removed from a packet via \emph{push} and \emph{pop}
actions. 
\petr{do we need the discussion of multiple tables/pipelining here? I still find it
  very confusing}

The actions associated with the packet are performed at the end of the pipeline.
If no rule matches the packet, the packet is dropped.

\liron{we should explain that: every packet is associated with an
  action set. This set is empty at the beginning and is changed
  according to the flow table matches. Each match is asociated with
  instructions that can add or modify actions in the action set and
  also to apply (now) all actions in the action set. By default action
  set is applied at the end of the pipeline.}

@@

\paragraph{Interaction between control and data planes: FlowMod commands}
The flow tables installed on the switches across the data plane
determine the network \emph{policy}.
Controllers can change the policy via sending  
control messages containing \emph{FlowMod commands}.
In a nutshell, a \emph{FlowMod} command either specifies a new flow entry or
a modification to an existing flow entry. \petr{The description below
  does not explain how a flow entry can be modified.}  
A flow entry consists of: (1)~a match 
field
%defined by a ternary pattern over packet header fields
(a value and a mask), (2)~a set of actions, and (3)~a priority.
%,a match and an action.

A FlowMod command can either add a flow entry or delete 
The standard processing of a FlowMod $\add$ command received by a switch is
as follows. 
If the switch already maintains a flow entry with \emph{exactly} the
same match and priority level, then the new flow entry will simply \emph{replace} it.
\petr{What if the two rules have different priorities?}
Otherwise, the flow entry will be installed in addition to existing
ones.
A $\dele$ command simply remove an existing flow entry with the same
match and priority  or does nothing if such an entry is not present.
\petr{Is this correct?}

@@

If rules with overlapping but not identical match fields have the same priority, 
inconsistencies can arise, since there is no clearly specified action
that should be applied to a packet matching the two entries.
In order to avoid such conflicts, a \emph{FlowMod} command can be equipped with a the $\checko$ flag:
if the switch maintains is a flow entry with an overlapping but not
identical match part with the same priority but a different action,
 and if the two rules have the same priority, then
\emph{FlowMod} will fail.
\petr{Is ``different action'' important here?}
The $\checko$ flag in a \emph{FlowMod} command helps resolving simple
conflicts between controllers, and will be instrumental
in implementing our synchronization abstractions.

%[[PK not sure this is the right place 
%implement synchronization objects which form
%the basis of our transactional interface.
%]]
We will use the following notation to create a \emph{FlowMod} message:
$\FlowMod(\match, \add/\dele, \action, \flags)$, where the first field
$\match$ denotes the match,
the second introduces 
$\action$ the action, and $\flags$ the flags to activate the $\checko$
feature.

\petr{Sometimes we also write something like op=add in flowmod, should
  we specify it here?}  
The \emph{FlowMod} message (and also the flow entries) include more parameters,
which however are irrelevant for our implementation: we suggest to
use the default values.

In order to read the current configuration of a switch (the existing flow entries), the controller should send
%By installing 
the OpenFlow \emph{add flow monitor} command with the \textsf{ofpt\_initial} optional flag.
%In reply the switch will send all the flow entries defined in the switch.
In this paper, we indicate sending this message and gathering the responds (the reply messages containing the flow entries) by \emph{read configuration} statement which also returns the configuration.
%, a controller can read the current configuration. 
% %(and subscribe for updates). 
%
%can i match multiple flows in multiple tables? yes, but only for FlowMod, not GroupMod: but that is good enough!%
\liron{I changed a bit the description of flow monitors following Petr's comment, I hope it's more clear now.}
\petr{Not very clear what reading the configuration means: is it done
  in an atomic way?}

%[[PK should go to the discussion?
%We also remark that our solution is only based on flow tables
%and does not make use
%of \emph{OpenFlow group tables}, nor do we make use of \emph{cookies}: rules
%are always identified by exact matches.
%]]

@@

Another important feature in OpenFlow is the \emph{bundle}. 
A controller can send multiple $\FlowMod$ commands  equipped with 
the same \emph{bundle identifier}.
These commands are buffered temporarily by
the switch until a \emph{bundle commit} message is received from the
controller. 
The bundled commands are then performed with \emph{all-or-nothing semantics}:
either all of the commands are performed, or none of them is. 
In particular, if a configuration request contained in one of the bundled commands cannot
be applied (e.g.,  all the command is rejected because of the set
$\checko$ flag and a conflicting flow entry), all  other commands in
the bundle are rejected and an error
message is sent to the controller that issued them.
%If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
The error message contains $\xid$, the identifier  first failed command in the
bundle, and $\ecode$, the \emph{error code} corresponding to the failure.   
%[[PK too detailed and not very clear
%Concretely, when the commit fails, the switch must generate an error message corresponding to the
%message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
%the error data field corresponding to that message and the error code corresponding to the failure. If
%multiple messages are generating errors, the switch may return only the first error found or generate
%multiple error messages for the same bundle.
%]]

The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, a bundle contains a \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

%When properly configured,
Bundle execution is considered controller atomic in the sense that all controllers can observe switch states (configurations) that either \emph{before} or \emph{after} the
bundle (i.e., all its commands) are executed. 
By specifying the flag \texttt{OFPBF\_ATOMIC},  
a bundle can also be  \emph{packet-atomic}, in a similar sense, which can prevent packets to being processed by intermediate (possibly invalid) configuration during a bundle.
In addition, by specifying the flag \texttt{OFPBF\_ORDERED},  bundle commands are executed according to the order they were added to the bundle, thereby allowing to perform dependent actions inside a bundle.
\liron{i added controller atomicity and order preservation, see that it make sense...}

@@

First, %if employed without any additional features, 
a bundle does not provide any means to modify the switch configuration 
based on a generic application-specific condition, depending on the
current configuration, except to limited types of constraints, e.g.,
or \emph{overlapping flow spaces} or insufficient free space.
Indeed, sometimes it is desirable to  define more generic conditions
under which configuration updates should or should not take place.

@@

Second, sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g., by supporting \emph{conditional modifications}.
Let us consider two examples.

@@

\petr{postpone the explanation of the right part of the figure until we come to CAS?}

@@

\subsection{Configuration space}

We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules), 
and the \emph{meta-configuration} that contains information used by the
controllers to synchronize their actions.

@@

\subsection{Synchronization}

Our framework provides the controllers with the ability to execute
sequences of \emph{operations} on the 
switch configuration in an \emph{atomic} (transactional) or \emph{non-atomic} mode.     
In the atomic mode, all operations the sequence (which we also refer
to as a \emph{transaction}) 
should either be executed in its entirety or completely rejected. 
In the non-atomic mode, no atomicity guarantees are provided. 
An operation here can be a regular FlowMod command or a synchronization
primitive, which, as we
show below, translates into a sequence of FlowMod commands operating on the
switch meta-configuration.

@@

\subsection{Synchronization primitives}

@@

As we will see, in the OpenFlow implementation of our mechanism, 
in addition to the ``normal'' configuration stored at the switch
(e.g., forwarding rules according to network policies), 
controllers will maintain additional information 
in the switch configuration: this information is used by the
controllers to synchronize. 
\petr{"As we will see,..." - looks funny compared with "As we will see ..." at the end of Sec 2}
\subsection{Synchronization Objects}\label{sec:t-if}

We implement transactional access to  two kinds of objects:
a read-write \emph{memory} abstraction
object, which we will use for storing the current used policy identifier, and a
\emph{resource-claimer} object, which allows controllers to claim a resource ID,
e.g., a policy \emph{identifier} in a
given (bounded or unbounded) set.
%While In theory one can use the memory abstraction to implement the resource claimer,
%however we found it far more simple and efficient to consider the resource claimer as base object.
%

With these base objects, we provide the following shared memory abstraction, with
write-compare access semantics:
\begin{itemize}
\item By performing $\textit{write}(\addr,k)$,
  $\addr,k\in\Nat$, the value at memory offset $\addr$ is set to $k$. For
  example, the offset $\addr$ may correspond to the policy identifier (\emph{pid}),
  but our approach is more general.

\item By performing an operation $\textit{\compare}(\addr,k)$, $\addr,k\in\Nat$,
a transaction can be aborted if the value at memory offset $\addr$ is not equal to $k$.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).
\end{itemize}

\petr{it is unclear why exactly we need these write and compare operations}

As we will see, the bundle feature is not required to implement these semantics.
Also note that memory read operations correspond to reading the switch
configuration, and are simple as they do not affect the execution of the transaction
(unlike the $\textit{\compare}$ operation). \petr{read operations "do not affect the execution of the transaction" appears strange, given that a transaction, if committed, "appears to execute sequentially". Something seems wrong here...}

Our resource-claimer object implements a weak type of locking, and exports operations \emph{claim},
\emph{unclaim} and \emph{\claimcheck} with the following sequential
semantics:
%The container synchronization object introduced in this paper
%is essentially a transactional memory abstraction which can be implemented
%in standard OpenFlow switches.
%Before presenting the technical details in the next section,%
%we give a brief overview.

%The postulated container object can store an arbitrary number of \emph{items}.
%As we will see, these items correspond to flow entries on the OpenFlow switch.
%The same item may even exist multiple times in the container (i.e., the container
%is a multi-set).
%The container
%provides two main functions:
\begin{itemize}
\item With $\textit{claim}(k)$,
  $k\in\Nat$, a controller $i\in[N]$ \emph{claims}
  identifier $k$.

\item With $\textit{unclaim}(k)$, $k\in\Nat$, controller $i\in[N]$
  \emph{unclaims} identifier $k$.

\item If $k$ is currently claimed by any controller,
$\textit{\claimcheck}(k)$, $k\in\Nat$,
aborts the current transaction.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).

@@

Again, none of these individual operations
requires bundles.

\petr{Would be great actually to motivate why we use reads, writes and claim ids. Maybe skecth a solution to some important problems using these primitives in the transactions.} \liron{we mention that its kind of locking. but I agree we can motivate it better}

@@

\petr{what is "offset" here (offset addr may correspond")?}\liron{I don't understand the comment}

\subsection{Implementation}\label{sec:t-impl}

We now describe how to implement our transactional
abstractions using standard OpenFlow features.

%features , including
%\emph{bundling}, \emph{conditional application}, and \emph{groups}.
%[[PK let us leave for the moment
%We will describe later how to implement some of these features
%in older OpenFlow versions where they are not
%supported. \textbf{TODO}
%]]
Algorithms~\ref{alg:claim}-\ref{alg:compare} describe implementations of
transactional operations  $\textit{claim}(x)$, $\textit{\claimcheck}(x)$,
$\textit{unclaim}(x)$, $\textit{write}(x)$ and $\textit{\compare}(x)$, respectively.
%Algorithm~\ref{alg:commit} describes the procedure executed by
%a transaction at \emph{commit time}.
Each operation produces OpenFlow command messages that should be sent to the switch, possibly as part of a bundle. 
\hide{ 
We describe each operation as a message factory \petr{did not get what is "message factory"} that returns OpenFlow commands which
should be sent to the switch as part of the same transaction.
We assume that every transaction has a unique identifier and is used to start a \emph{bundle}.
All messages generated during a transaction invocation, are wrapped by bundle messages carrying
the bundle's ID and followed by bundle commit message.
Messages with same bundle ID are considered to belong to same bundle and are performed atomically.
\liron{I don't think we need all of this bundle staff here. we need to move it to next sextion - policy updates}
}
%The OpenFlow messages that we use (besides )

%Our implementation is based on the \emph{FlowMod} command. 
To simplify the algorithms we set the FlowMod command field values with integers, according to the following guidelines:
\begin{itemize}

@@

In addition we use "op" as shortened form to a field that indicates whether the command should $\add$ or $\dele$ the specified flow entry. This corresponds to the FlowMod field named "command" which can receive values OFPFC\_ADD and OFPFC\_DELETE.

\paragraph{The resource claimer}
\petr{
the "resource claimed" part a hard read, the trick of using s and values is smart, deserves to be explained better
}

\petr{Alg 2 returns "cmd1, cmd2", what does it mean? Isn't it supposed to return a boolean?}
\petr{what is this "$\concat$" operation? simple concatenation?}
\petr{sometimes FlowMod commands have flag and action, sometimes not, sometimes flag=0, semantics of all this is unclear.}
\liron{There is now some text above that should help to understand this part (the resource claimer) and the next one. I hope it does.}

As can be seen in Algorithms \ref{alg:claim}-\ref{alg:unclaim}, for the resource-claimer,
the match part of the added flow entry represents the claimed ID (the ID is concatenated twice). 
In addition the mask part of the match (controller ID concatenated with all-ones) varies from one controller
to another in order to allow multiple controllers to claim the same number without overriding the same flow entry. 
We can see that for any two resource IDs $x_1$ and $x_2$ and controller IDs $c_1\neq c_2$,
the match patterns $m_1=(x_1\concat x_1, c_1\concat all-1)$ and $ m_2=(x_2\concat x_2, c_2\concat all-1)$ are not equal,
even if $x_1=x_2$; therefore, they can both be used to claim and unclaim identifiers without affecting each other.

In order to check if a number is claimed, we try to create the
corresponding entry while setting the
flag \texttt{\checko}, and using different actions ($2$ instead of $1$), thereby inflicting
a failure in case an entry with overlapping match value exists.
If the check succeeds, we delete the entry. Considering $m_1$ and $m_2$ as defined above,
and their respective actions $a_1=x_1\concat c1$ and $a_2=x_2\concat c2$,  they overlap
with different actions iff $x_1=x_2$. Therefore, if one is used in \claimcheck after the other used in the claim operation,
a conflict is detected iff $x_1=x_2$, as desired.

@@

\begin{algorithm}[H]
    \caption{$\textit{claim}(x)$}
    \label{alg:claim}
    \begin{algorithmic}[1]
    \Require \emph{self} as the calling controller id.
%    \Ensure
    		\State $value \gets x\concat x$
    		\State $mask \gets self\concat all-1$
	    	\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets 1$
    		\State $flag \gets 0$
    		\State $cmd\gets \emph{FlowMod}(match, op = \add, flag, action) $
			\Return $cmd$
    \end{algorithmic}
\end{algorithm}

>>>>

\newcommand{\hide}[1]{}
\newcommand{\concat}[0]{\cdot}
\newcommand{\append}[0]{\frown}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\cas}{CAS\xspace}
\newcommand{\claimcheck}{check\xspace}
\newcommand{\compare}{compare\xspace}
\newcommand{\memwrite}{write\xspace}

@@

%\title{Solving Consensus with OpenFlow}

@@

%[[PK sorry for stubborness :)
\title{On the Synchronization Power of the OpenFlow Data Plane}
%\\ {\Large Towards Atomic Transactions In-Band}}
%]]

@@

We show that powerful synchronization constructs can be implemented
directly \emph{in OpenFlow}, i.e., using the \emph{de facto} standard SDN
protocol. 
In our implementation, a part of the data-plane configuration space is
used as a shared memory storing information about  contention and
conflicts between distributed controllers.
A transaction can then contain standard OpenFlow control operation
as well as  \emph{synchronization primitives} operating on this shared
memory. 
Our synchronization primitives allow the controllers to define complex notions
of conflicts that can be used for coordinating configuration updates that
operate over \emph{independent flow spaces}, but where dependencies are introduced
in the mapping of the flows to the physical infrastructure. This is
important, e.g., for load-balancing applications.

@@

More precisely, the match part of a rule is 
a ternary pattern over packet \emph{header} fields.
The pattern is represented as a \emph{value} and a \emph{mask}. 
In the mask, certain header fields, e.g., port numbers or destination and source
addresses, can be \emph{wildcarded}, stipulating that their content does
not affect the match. 
For examples, \texttt{src=10.1.14.0, ff.ff.ff.f0} matches all packets
with source addresses in the range \texttt{10.1.14.0}--\texttt{10.1.14.127}.
Certain field, such as \emph{metadata}, can be arbitrarily
bitmasked. 
In our implementations, we are going to use bitmasking of the metadata
filed for performing conditional operations on  its content.   
\petr{Example?}

A \emph{configuration} of an OpenFlow switch is represented as a
collection \emph{flow table}.
A flow table is a set of \emph{flow entries}, each containing a rule,
with a match, an action, and a \emph{priority} level, and flow entries
are ordered according to their priorities.  
In the simplest setting, a switch maintains one flow table (table 0).  
A data-plane packet arriving at a switch is first checked against
the rule with the highest priority and in table $0$. 
If the header of the packet fits the match fields of that rule, 
a default \emph{instruction} associates the packet with the corresponding action.
Otherwise, the packet is checked against flow  entries with lower
priorities, and if no matching rule is found, the packet is dropped.

%[[PK move pipelining to the discussion?
\hide{
Furthermore, an OpenFlow switch may maintain multiple flow tables
forming a \emph{pipeline}.
Data-plane packets are checked against flow entries in the pipelined
tables to determine the set of actions the switch should perform on the packet.      
A data-plane packet arriving at a switch is first checked against
the rule with the highest priority and in table $0$. 
If the header of the packet fits the match fields of that rule, 
a default \emph{instruction} associates the packet the corresponding action.
Otherwise, the packet is checked against flow  entries with lower
priorities.  
Specific instructions can be used to define
additional tables to be visited (\textit{goto} instructions), to
modify the set of actions associated with the packet (by adding,
deleting or modifying actions), or immediately apply some actions in the action set. 
During the traversal of the flow tables, the packet can also be equipped with \emph{meta-tags} to
transfer temporary information between tables.
The meta-data field is part of the header that can be
inserted or removed from a packet via \emph{push} and \emph{pop}
actions.

The actions associated with the packet are performed at the end of the pipeline.
If no rule matches the packet, the packet is dropped.
\petr{do we need the discussion of multiple tables/pipelining here? I still find it
  very confusing}

\liron{we should explain that: every packet is associated with an
  action set. This set is empty at the beginning and is changed
  according to the flow table matches. Each match is asociated with
  instructions that can add or modify actions in the action set and
  also to apply (now) all actions in the action set. By default action
  set is applied at the end of the pipeline.}
}

@@

\paragraph{Interaction between the planes: FlowMod commands}
The flow tables installed on the switches across the data plane
determine the network \emph{policy}.
Controllers can change the policy via sending  
control messages containing \emph{FlowMod commands}.
In a nutshell, a \emph{FlowMod} command either specifies a new flow entry or
a modification to an existing flow entry.  
%[[PK already said
%A flow entry consists of: (1)~a match 
%field
%(a value and a mask), (2)~a set of actions, and (3)~a priority.
%]]

A FlowMod command can either add a flow entry or delete a flow entry.
The standard processing of a FlowMod $\add$ command received by a switch is
as follows. 
If the switch already maintains a flow entry with \emph{exactly} the
same match and priority level, then the new flow entry will simply \emph{replace} it.
Otherwise, the flow entry will be installed in addition to existing
ones.
A $\dele$ command simply remove an existing flow entry with the same
match and priority  or does nothing if such an entry is not present.
\petr{Is this correct?}

@@

If rules with overlapping but not identical match fields have the same priority, 
inconsistencies can arise, since there is no clearly specified action
that should be applied to a packet matching the two entries.
In order to avoid such conflicts, a \emph{FlowMod} command can be equipped with a the $\checko$ flag:
if the switch maintains is a flow entry with an overlapping but not
identical match part with the same priority but a different action,
 and if the two rules have the same priority, then
\emph{FlowMod} will fail.
The $\checko$ flag in a \emph{FlowMod} command helps resolving simple
conflicts between controllers, and will be instrumental
in implementing our synchronization abstractions.

%[[PK not sure this is the right place 
%implement synchronization objects which form
%the basis of our transactional interface.
%]]
We will use the following notation to create a \emph{FlowMod} message:
$\FlowMod(\match, op, \action, \flags)$, where 
%$\match$ denotes the match,
$op$ is $\add$ or $\dele$, and $\flags$ are used, e.g.,  to activate the $\checko$
feature.
A standard \emph{FlowMod} message (and also the flow entries) may specify
more parameters, which are however irrelevant for our implementation, we suggest to
use the default values.

In order to read the current configuration of a switch (the existing flow entries), the controller should send
the OpenFlow \emph{add flow monitor} command with the \textsf{ofpt\_initial} optional flag.
In this paper, we indicate sending this message and gathering the
responds (the reply messages containing the flow entries) 
by \textit{read-config} statement which returns the configuration.

\hide{
%[[PK should go to the discussion?
We also remark that our solution is only based on flow tables
and does not make use
of \emph{OpenFlow group tables}, nor do we make use of \emph{cookies}: rules
are always identified by exact matches.
%]]
}

@@

Another important feature in OpenFlow is \emph{bundling}. 
A controller can send multiple $\FlowMod$ commands  equipped with 
the same \emph{bundle identifier}.
These commands are buffered temporarily by
the switch until a \emph{bundle commit} message is received from the
controller. 
The bundled commands are then performed with \emph{all-or-nothing semantics}:
either all of them are performed, or none of them is. 
In particular, if a configuration request contained in one of the bundled commands cannot
be applied (e.g., the command is rejected because of the set
$\checko$ flag and a conflicting flow entry), all  other commands in
the bundle are rejected and an error
message is sent to the controller that issued them.
%If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle,
The error message contains $\xid$, the identifier of the first failed command in the
bundle, and $\ecode$, the \emph{error code} corresponding to the failure.

A bundle begins with a  \texttt{ofpt\_bundle\_control} message of type
\texttt{ofpbct\_open\_request} (creating the bundle), wraps
each of its \emph{FlowMod} command in a message  
\texttt{ofpt\_bundle\_add\_message} equipped with the \emph{bundle
  identifier}, and ends with a \texttt{ofpt\_bundle\_control} message of
type \texttt{ofpbct\_commit\_request} (committing the bundle).

%[[PK too detailed and not very clear
%Concretely, when the commit fails, the switch must generate an error message corresponding to the
%message that could not be applied. The error message must use the identifier (\emph{xid}) of the offending message,
%the error data field corresponding to that message and the error code corresponding to the failure. If
%multiple messages are generating errors, the switch may return only the first error found or generate
%multiple error messages for the same bundle.
%]]

%Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

%When properly configured,
Bundle execution is considered \emph{controller-atomic} 
in the sense that all controllers can observe switch states (configurations) that either \emph{before} or \emph{after} the
bundle (i.e., all its commands) are executed. 
By specifying the flag \texttt{ofpbf\_atomic},  
a bundle can also be  \emph{packet-atomic}, in a similar sense, which can prevent packets to being processed by intermediate (possibly invalid) configuration during a bundle.
In addition, by specifying the flag \texttt{ofpbf\_ordered},  bundle commands are executed according to the order they were added to the bundle, thereby allowing to perform dependent actions inside a bundle.
%\liron{i added controller atomicity and order preservation, see that it make sense...}

@@

First, %if employed without any additional features, 
a bundle does not provide any means to modify the switch configuration 
based on a generic application-specific condition, depending on the
current configuration, except to limited types of constraints, such as
overlapping flow spaces or insufficient free space.
But we might need more generic conditions under which configuration
updates 
should or should not take place.

@@

Second, sometimes it is desirable to react to conflicts in smarter ways than
by simply aborting an operation, e.g., by supporting \emph{conditional modifications}.
Let us consider two examples.

@@

%\petr{postpone the explanation of the right part of the figure until we come to CAS?}

@@

\paragraph{Configuration space}
%
We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules), 
and the \emph{meta-configuration} that contains information used by the
controllers to synchronize their actions.
Flow entries in the meta-configuration can be referred to as shared
memory locations, provided a total addressing scheme.

@@

\paragraph{Synchronization}
 %
Our framework provides the controllers with the ability to execute
sequences of \emph{operations} on the 
switch configuration in an \emph{atomic} (transactional) or \emph{non-atomic} mode.     
In the atomic mode, all operations the sequence (which we also refer
to as a \emph{transaction}) 
should either be executed in its entirety or completely rejected. 
In the non-atomic mode, no atomicity guarantees are provided. 
An operation here can be a regular FlowMod command or a synchronization
primitive, which, as we
show below, translates into a sequence of FlowMod commands operating on the
switch meta-configuration.

@@

\paragraph{Synchronization primitives}
%
\hide{

@@

As we will see, in the OpenFlow implementation of our mechanism, 
in addition to the ``normal'' configuration stored at the switch
(e.g., forwarding rules according to network policies), 
controllers will maintain additional information 
in the switch configuration: this information is used by the
controllers to synchronize. 
\petr{"As we will see,..." - looks funny compared with "As we will see ..." at the end of Sec 2}
\subsection{Synchronization Objects}\label{sec:t-if}
}

Following the work on consistent policy
updates~\cite{network-update,stn}, we focus on the consistent use of  \emph{policy
identifiers}.
Intuitively, a controller should 
equip each policy update it wants to install with a distinct
identifier
%[[PK do not want to say ``unique'': we also need to motivate claimer-id]] 
and the installation should succeed only if the identifier of the
currently installed policy has not changed since the last time the
controller read it.

Therefore, we provide the following transactional operations.
A $\textit{write}(\addr,k)$ operation sets the content at memory location
$\addr$ to $k$. 
%
%For  example, the offset $\addr$ may correspond to the policy identifier (\emph{pid}),
%  but our approach is more general.
%
A $\textit{\compare}(\addr,k)$ operation \emph{aborts} the transaction
if and only if the content at
$\addr$ is \emph{not} $k$.
Intuitively, by combining compare and write with a set of regular
FlowMod commands in a transaction we can make sure that the
configuration is changed and the current policy identifier is updated
only if the check operation \emph{succeeds} (does not abort).

To make sure that policy identifiers do not grow without bound~\cite{stn}, we
also provide a \emph{id-claimer} object.
The object implements a weak type
of locking by allowing a controller to claim a \emph{resource identifier} in a
given set, release an identifier,  and check if a given identifier
is currently claimed by any controller.
%While In theory one can use the memory abstraction to implement the resource claimer,
%however we found it far more simple and efficient to consider the resource claimer as base object.
%
More precisely, the id-claimer object exports three operations \emph{claim},
\emph{unclaim} and \emph{\claimcheck} with the following sequential
semantics:
%The container synchronization object introduced in this paper
%is essentially a transactional memory abstraction which can be implemented
%in standard OpenFlow switches.
%Before presenting the technical details in the next section,%
%we give a brief overview.

%The postulated container object can store an arbitrary number of \emph{items}.
%As we will see, these items correspond to flow entries on the OpenFlow switch.
%The same item may even exist multiple times in the container (i.e., the container
%is a multi-set).
%The container
%provides two main functions:
\begin{itemize}
\item With $\textit{claim}(k)$,
%  $k\in\Nat$, 
a controller %$i\in[N]$ 
\emph{claims}  identifier $k$.

\item With $\textit{unclaim}(k)$, %$k\in\Nat$, 
the controller %$i\in[N]$
  \emph{unclaims} identifier $k$.
An identifier $k$ is called \emph{claimed} at a given point of an
execution if there is a controller that has performed $\textit{claim}(k)$
but has not yet performed  $\textit{unclaim}(k)$.

\item If $k$ is currently claimed by any controller,
$\textit{\claimcheck}(k)$ %, $k\in\Nat$,
aborts the current transaction.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).

@@

%As we will see, the bundle feature is not required to implement these semantics.
%Also note that memory read operations correspond to reading the switch
%configuration, and are simple as they do not affect the execution of the transaction
%(unlike the $\textit{\compare}$ operation). \petr{read operations "do not affect the execution of the transaction" appears strange, given that a transaction, if committed, "appears to execute sequentially". Something seems wrong here...}

@@

\paragraph{Implementation}\label{sec:t-impl}
%
We now describe how to implement our transactional
abstractions using standard OpenFlow features.
Essentially, we translate each of the transactional operations into a
sequence of FlowMod commands.

The implementation of $\exec\{op_1,\ldots,op_k\}$ and
$\execatomic\{op_1,\ldots,op_k\}$ constructs
invoked by a control application is simple.  
First we translate  $op_1,\ldots,op_k$ into a sequence of FlowMod
commands. If $op_i$ is already a FlowMid command, the translation is 
trivial. If $op_i$ is a synchronization primitive in
$\{\textit{claim},  \textit{checkclaim}, \textit{unclaim},
\textit{write}, \textit{compare}\}$, we use
Algorithms~\ref{alg:claim}, \ref{alg:check}, \ref{alg:unclaim},
\ref{alg:write}, and \ref{alg:compare}, respectively, to get the
corresponding command(s).

To run  $\exec\{op_1,\ldots,op_k\}$, we simply send the commands in
the resulting sequence, one by one, to the switch. 
%
To run  $\execatomic\{op_1,\ldots,op_k\}$, we 
additionally create a bundle, wrap each of the resulting commands in a
bundle message with the corresponding bundle identifier, complete it
with a bundle commit message.
The we wait until the
corresponding responses are received. 
If an error message is received
it is returned to the control application,
otherwise $\ack$ is returned.

We now describe Algorithms~\ref{alg:claim}-\ref{alg:write} in more
detail. 
Each of the operations generates one or two FlowMod commands. 
The \textit{match} part in these commands specifies only the $64$bit meta-data
field in the header space, leaving the remaining fields in default
values.  
In defining the value and the mask of the math part, we 
use notation $a\concat b$ to represent a concatenation of two
$32$bit integers $a$ and $b$: $a\concat b := (a<<32)+b$.
We also write \texttt{*} for  a $32$bit 
string of $1$'s used in a mask.  
%
The action part in these commands is defined as an integer, which in
practice can be implemented by a \emph{set meta-data} instruction where the written value is that integer.

\hide{
%[[PK too lengthy, partially said before
To simplify the presentation we set the FlowMod command field values with integers, according to the following guidelines:
\begin{itemize}

@@

In addition we use "op" as shortened form to a field that indicates whether the command should $\add$ or $\dele$ the specified flow entry. This corresponds to the FlowMod field named "command" which can receive values \textt{ofpfc\_add} and \texttt{ofpfc\_delete}.
%]]
}

\paragraph{The id-claimer}

In Algorithms \ref{alg:claim}-\ref{alg:unclaim}, 
the match part of the constructed commands represents a concatenation
of two claimed identifiers. 
In addition, every controller uses a unique mask part of the match
defined as a distinct $32$bit controller identifier concatenated with
\texttt{*}.
This allows multiple controllers to claim the same number without
overriding the same flow entry.
%
For example, for any two resource identifiers $x_1$ and $x_2$ and controller identifiers $c_1\neq c_2$,
the match patterns $m_1=(x_1\concat x_1, c_1\concat \texttt{*})$ and $ m_2=(x_2\concat x_2, c_2\concat \texttt{*})$ are distinct,
even if $x_1=x_2$. 
Therefore, they can both be used to claim and unclaim identifiers without affecting each other.

In order to check if an identifier is claimed, a controller tries to add the
a new entry setting the flag \texttt{\checko}, and using a different
action ($2$ instead of $1$), 
thereby inflicting a failure in case an entry with overlapping match value exists.
If the check succeeds, we delete the entry. 
In the example above above,
and their respective actions $a_1=x_1\concat c1$ and $a_2=x_2\concat c2$,  they overlap
with different actions iff $x_1=x_2$. Therefore, if one is used in \claimcheck after the other used in the claim operation,
a conflict is detected iff $x_1=x_2$, as desired.

@@

\begin{algorithm}[H]
    \caption{$\textit{claim}(x)$}
    \label{alg:claim}
    \begin{algorithmic}[1]
    \Require \emph{self} as the calling controller id.
%    \Ensure
    		\State $value \gets x\concat x$
    		\State $mask \gets self\concat \texttt{*}$
	    	\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets 1$
    		\State $flag \gets 0$
    		\State $cmd\gets \emph{FlowMod}(match, op = \add, flag, action) $
			\Return $cmd$
    \end{algorithmic}
\end{algorithm}
